---
title: "Regression Models - Course Project"
author: "Pier Lorenzo Paracchini"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
```{r settingsAndDependencies, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Sys.setlocale("LC_ALL", "C")
require("datasets")
data("mtcars")
```  
#Executive Summary

#Data Analysis
##Exploratory Data Analysis
The dataset include `r dim(mtcars)[1]` observations of `r dim(mtcars)[2]` features. Each onservations comprises information for a specific automobiles (1973 - 1974 model). More information about the features can be found running `help(mtcars)`.  

Focus is on two specific features __mpg__ (miles per US gallon) and __am__ (type of transmission - __0: automatic__, __1: manual__). The available datase contains `r table(mtcars$am)[[1]][1]` automatic car models and `r table(mtcars$am)[[2]][1]` manual car models. The observations of `mpg` by type of transmission `am` can be seen in the hystograms below (__note!__ black line represents the sample mean for the group).

```{r ref.label="histoMpgByAm", echo=FALSE, warning=FALSE, message=FALSE, fig.height=3, fig.width=6}
```  

The `am` feature (predictor) is visibly related to the `mpg` feature (outcome) as we can see from the sample mean of each group. The "automatic" group has a lower sample mean (`r round(mpg_avg_0,3)` Miles/ Gallon) than the "manual" group (`r round(mpg_avg_1,3)` Miles/ Gallon).  

##The Regression Model
Let's fit a __linear model__ using `mpg` (as outcome) with `am` as predictor with errors $\epsilon_i \stackrel{iid}{\sim} N(0, \sigma^2)$. The coefficients of the models are     
```{r echo=FALSE}
mpg <- mtcars$mpg; am <- as.factor(mtcars$am)
fit <- lm(mpg ~ am) 
summary(fit)$coefficient
```  
```{r echo=FALSE}
par(mfrow = c(2,2))
plot(fit)
```
From the __Residual vs Fitted__ plot we can see that the model is just able to predict two possible values for the estimated `mpg` (mile per gallon) based on the predictors - __17.147 for automatic models__ and __24.392 for manual models__ with an estimated residuals standard error of __`r round(summary(fit)$sigma,3)`__ (mile per gallon). The __QQplot__ diagnoses normality of the errors. 

##Is an automatic or manual transmission better for MPG?  
Looking at the coefficients   
```{r echo=FALSE}
coeff_fit <- summary(fit)$coefficient
ci_95_beta0 <- coeff_fit[1,1] + c(-1,1) * qt(.975, df = fit$df) * coeff_fit[1,2] 
ci_95_beta1 <- coeff_fit[2,1] + c(-1,1) * qt(.975, df = fit$df) * coeff_fit[2,2] 
coeff_fit
```  

* __automatic__ models (reference group) have an estimated mpg of `r  round(coeff_fit[1,1],3)` (miles per gallon) with a standard error of `r  round(coeff_fit[1,2],3)` (miles per gallon).

* __manual__ models have an increased estimated mpg of `r  round(coeff_fit[2,1],3)` (miles per gallon) (over the reference group) with a standard error of `r  round(coeff_fit[2,2],3)` (miles per gallon). The P-value of `r  round(coeff_fit[2,4],3)` (statistically significant) confirms that the __null hypothersis__ (having an increase/ decrease over the reference group null) can be rejected.

According to this very simple linear model, __automatic__ transmission is better for __mpg__ than __manual__ transmission. 

##Quantify the MPG difference between automatic and manual transmissions  
Based on the linear model previosly created we can state that 

*  __automatic__ models use an estimated mpg of `r  round(coeff_fit[1,1],3)` (miles per gallon) with a 95% confidence interval included in [`r round(ci_95_beta0[1],3)` - `r round(ci_95_beta0[2],3)`] miles per gallon.
*  __manual__ models use an increased estimated mpg of `r  round(coeff_fit[2,1],3)` (miles per gallon) over the reference group with a 95% confidence interval included in [`r round(ci_95_beta1[1],3)` - `r round(ci_95_beta1[2],3)`] miles per gallon.


#Description __(ToBeDeleted)__  
__Interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome)__. They are particularly interested in the following two questions:

* “Is an automatic or manual transmission better for MPG”  
* "Quantify the MPG difference between automatic and manual transmissions"

__Constraints__ Report must be:

* Written as a PDF printout of a compiled (using knitr) R markdown document.  
* Brief. Roughly the equivalent of 2 pages or less for the main text. Supporting figures in an appendix can be included up to 5 total pages including the 2 for the main report. The appendix can only include figures.  
* Include a first paragraph executive summary.


#Appendix
Code chunks and plots used in the previous sections.  
##Dataset Structure    
```{r dataStructure, echo=TRUE, warning=FALSE, message=FALSE}
str(mtcars)
```
### Some more details about `mpg` and `am` features  
Some `mpg` relevat statistics.  
```{r mpgDetails, echo=TRUE, eval=TRUE, collapse=TRUE}
summary(mtcars$mpg)
```  
Number of car models by type of transmission.  
```{r amDetails, echo=TRUE, eval=TRUE, collapse=TRUE}
table(mtcars$am)
```  
Plot showing the number of car models by type of transmission and mpg.  
```{r amDetailsExtra, echo=TRUE, eval=TRUE, collapse=TRUE}
require(dplyr)
require(ggplot2)
freqData <- as.data.frame(table(mtcars$mpg, mtcars$am))
names(freqData) <- c("mpg", "am", "freq")

g <- ggplot(filter(freqData, freq > 0), aes(x = am, y = mpg))
g <- g  + scale_size(guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+2, show_guide = TRUE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")
g <- g + ggtitle("Mile Per (US) Gallon vs. Transmission Type")
g <- g + xlab("Transmission Type - 0: automatic, 1: manual")
g <- g + ylab(" Miles/(US) Gallon")
g
```  
Histograms for `mpg` by `am` (transmission type)  
```{r histoMpgByAm, echo=TRUE, eval=FALSE, warning=FALSE, message=FALSE}
require(ggplot2)
require(dplyr)

mpg_avg_0 <- mean(filter(mtcars, am == 0)$mpg)
mpg_avg_1 <- mean(filter(mtcars, am == 1)$mpg)


#Other possibility
data_0 <- data.frame(mpg = mtcars$mpg, am = as.factor(mtcars$am))
data_1 <- data.frame(am = c("0", "1"), mean = c(mpg_avg_0, mpg_avg_1))

g <- ggplot(data_0, aes(x = mpg, fill = am))
g <- g + geom_histogram(colour = "black", binwidth=1)
g <- g + facet_grid(am ~ .)
g <- g + geom_vline(data = data_1, aes(xintercept = mean), lwd = 1)
g

```  
