---
title: "Regression Models - Course Project"
author: "Pier Lorenzo Paracchini, 18.12.2015"
output:
  word_document: default
  pdf_document: default
  html_document:
    keep_md: yes
---
```{r settingsAndDependencies, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Sys.setlocale("LC_ALL", "C")
require(dplyr)
require("datasets")
data("mtcars")

mpg <- mtcars$mpg
am <- as.factor(mtcars$am)
hp <- mtcars$hp

mtcars_0 <- filter(mtcars, am == 0)
mtcars_1 <- filter(mtcars, am == 1)

mpg_avg_0 <- mean(mtcars_0$mpg)
mpg_avg_1 <- mean(mtcars_1$mpg)

fit_a <- lm(mpg ~ am)
coeff_fita <- summary(fit_a)$coefficient
cia_95_beta0 <- coeff_fita[1,1] + c(-1,1) * qt(.975, df = fit_a$df) * coeff_fita[1,2] 
cia_95_beta1 <- coeff_fita[2,1] + c(-1,1) * qt(.975, df = fit_a$df) * coeff_fita[2,2] 

fit_b <- lm(mpg ~ am + hp)
coeff_fitb <- summary(fit_b)$coefficient
cib_95_beta0 <- coeff_fitb[1,1] + c(-1,1) * qt(.975, df = fit_b$df) * coeff_fitb[1,2] 
cib_95_beta1 <- coeff_fitb[2,1] + c(-1,1) * qt(.975, df = fit_b$df) * coeff_fitb[2,2] 

```  
#Executive Summary
__Automatic__ models are better for `mpg`. Actually __automatic__ models have an estimated mpg of `r  round(coeff_fitb[1,1],3)` (miles per gallon)  with a 95% confidence interval included in [`r round(cib_95_beta0[1],3)` - `r round(cib_95_beta0[2],3)`], while __manual__ models have an estimated increased mpg of `r  round(coeff_fitb[2,1],3)` (miles per gallon) over the __automatic__.

#Data Exploration and Analysis
The `mtcars` dataset include `r dim(mtcars)[1]` observations of `r dim(mtcars)[2]` features. Each onservations comprises some information for specific automobiles (1973 - 1974 models). A scatterplot matrix for all features available in the datase is available in __Appendix, Figure 1__. The focus is on `mpg` (miles per gallon) and `am` (type of transmission - 0: automatic, 1: manual). 

`mtcars` dataset contains `r table(mtcars$am)[[1]][1]` automatic car models and `r table(mtcars$am)[[2]][1]` manual car models. The observed `mpg` by type of transmission `am` can be seen in the provided hystogram (see __Appendix, Figure 2__). From the sample data, the `am` (predictor) is visibly related to `mpg` (outcome) as we can see from the sample mean of each group. Specifically the "automatic" group has a lower sample mean (`r round(mpg_avg_0,3)` miles per gallon) than the "manual" group (`r round(mpg_avg_1,3)` miles per gallon).

#Regression Model

##Simple Model
The simple model uses `mpg` (as outcome) using `am` as the only predictor (__`mpg ~ am`__). The coefficients of the fitted model are     
```{r echo=FALSE}
summary(fit_a)$coefficient
```  
From the Residual vs Fitted plot (see __Appendix, Figure 3__) we can see that the model is just able to predict two possible values for the estimated `mpg` based on the value of the predictor - __`r  round(coeff_fita[1,1],3)` mpg for automatic models__ and __`r  round(coeff_fita[1,1],3) + round(coeff_fita[2,1],3)` mpg for manual models__ with an estimated residuals standard error of __`r round(summary(fit_a)$sigma,3)` mpg__. __`R- squared`__ indicates that this model is able to explain only __`r round(summary(fit_a)$r.squared * 100,2)`%__ of the total variability of the data. The Q-Q plot (see __Appendix, Figure 3__) confirms the normality of the errors (assumption).

A more detailed interpration of the simple model can be found in __Appendix, Simple Model Interpretation__.

##Extending the Model adding new features  
The simple model is quite limited and other available features could be used to identify a "better" model. A possible feature that may be valuable investigating is `hp` (__gross horse power__ in hp). The relationship between `mpg` vs. `hp` by `am` can be seen in __Appendix, Figure 4__.   

Using the __nested model testing technique__ for the following nested models we can see that, __added feature of model 2 are necessary over model 1__ (P-value < 0.05).

```{r nestedModelResults, echo=FALSE}
anova(fit_a, fit_b)
```  
Model 2 (__`mpg ~ am + hp`__) seems to provide a better "description" of the response and it will be used to answer the original questions. See the Residual vs Fitted plot and Q-Q plot (see __Appendix, Figure 5__). The estimated residuals standard error of __`r round(summary(fit_b)$sigma,3)` mpg__ is decreased and __`R- squared`__ of __`r round(summary(fit_b)$r.squared * 100,2)`%__ is increased (compared to model 1).

__Note!!__ The same process can be executed adding other features to the model in a nested fashion - investigating the overall effect of the new features on the model in an incremental way.

##Findings & Interpretation
Model 2 (__`mpg ~ am + hp`__) has the following coefficients  
```{r echo=FALSE}
coeff_fitb
```  

* __automatic__ models (reference group - `am = 0`) have an estimated mpg of `r  round(coeff_fitb[1,1],3)` (miles per gallon) with a standard error of `r  round(coeff_fitb[1,2],3)` (miles per gallon).

* __manual__ models (`am = 1`) have an increased estimated mpg of `r  round(coeff_fitb[2,1],3)` (miles per gallon) (over the reference group) with a standard error of `r  round(coeff_fitb[2,2],3)` (miles per gallon). The P-value of `r  round(coeff_fitb[2,4],3)` is statistically significant, reject the __null hypothersis__ (having an increase/ decrease over the reference group null).

According to this very simple linear model, __automatic__ transmission is better for __mpg__ than __manual__ transmission. 

Based on the linear model previosly created we can state that 

*  __automatic__ models use an estimated mpg of `r  round(coeff_fitb[1,1],3)` (miles per gallon) with a 95% confidence interval included in [`r round(cib_95_beta0[1],3)` - `r round(cib_95_beta0[2],3)`] miles per gallon.
*  __manual__ models use an increased estimated mpg of `r  round(coeff_fitb[2,1],3)` (miles per gallon) over the reference group with a 95% confidence interval included in [`r round(cib_95_beta1[1],3)` - `r round(cib_95_beta1[2],3)`] miles per gallon.

\newpage

#Appendix  
##Figure1: Scatterplot Matrix for features in `mtcars` dataset  
```{r scatterplotMatrix, echo=FALSE}
plot(mtcars)
```  

##Figure2: Hystogram of observed `mpg` by `am` (type of transmission)  
Note the _blue_ line represent the sample mean for each of the type of transmission (0: automatic, 1: manual).  
```{r histoMpgByAm, echo=FALSE, warning=FALSE, message=FALSE, fig.height=3, fig.width=6}
require(ggplot2)
require(dplyr)

data_0 <- data.frame(mpg = mpg, am = as.factor(am))
data_1 <- data.frame(am = c("0", "1"), mean = c(mpg_avg_0, mpg_avg_1))

g <- ggplot(data_0, aes(x = mpg, fill = am)) +  theme(text = element_text(size=8))
g <- g + geom_histogram(colour = "black", binwidth=1)
g <- g + facet_grid(. ~ am)
g <- g + geom_vline(data = data_1, aes(xintercept = mean), lwd = 1, col="blue")
g

```  

##Figure3: Residual Plots for (_`mpg ~ am`_) model

```{r resPlotsMpgVsAm, echo=FALSE, fig.height=3, fig.width=6}
par(mfrow = c(1,2), cex=0.7)
plot(fit_a, which = c(1,2))
```  

##Figure4: (_`mpg vs. hp`_) by _`am`_ (transmission type) plots 
For each type of transmission the __unadjusted__ line (black), __adjusted__ lines (lightblue for automatic, salmon for manual) and sample averages (horizontal) lines (lightblue for automatic, salmon for manual) are plot.   
```{r otherFeaturesPlotsByAm, echo=FALSE, fig.height=3, fig.width=6}
par(mfrow = c(1,1), cex=0.7)

plot(hp, mpg, type="n")
abline(lm(mpg ~ hp), lwd = 2)
fit <- lm(mpg ~ hp + am)
abline(coef(fit)[1], coef(fit)[2], lwd = 2, col="lightblue" )
abline(h = mean(mtcars_0$mpg), lwd = 2, col="lightblue")
abline(coef(fit)[1] + coef(fit)[3], coef(fit)[2], lwd = 2, col="salmon")
abline(h = mean(mtcars_1$mpg), lwd = 2, col="salmon")
points(mtcars_0$hp, mtcars_0$mpg, pch = 21, col = "black", bg = "lightblue", cex = 2)
points(mtcars_1$hp, mtcars_1$mpg, pch = 21, col = "black", bg = "salmon", cex = 2)
legend("topright", legend = c("0 - Automatic", "1 - Manual"), col = c("lightblue", "salmon"), fill = c("lightblue", "salmon"))
title("mpg vs. hp - adjusted by am")
```  

##Figure5: Residual Plots for (_`mpg ~ am + hp`_) model  
```{r resPlotsMpgVsAmWtHp, echo=FALSE, fig.height=3, fig.width=6}
par(mfrow = c(1,2), cex=0.7)
plot(fit_b, which = c(1,2))
```  

##Simple Model Interpretation
Looking at the coefficients of the simple model (`mpg ~ am`)  
```{r echo=FALSE}
coeff_fita
```  

* __automatic__ models (reference group) have an estimated mpg of `r  round(coeff_fita[1,1],3)` (miles per gallon) with a standard error of `r  round(coeff_fita[1,2],3)` (miles per gallon).

* __manual__ models have an increased estimated mpg of `r  round(coeff_fita[2,1],3)` (miles per gallon) (over the reference group) with a standard error of `r  round(coeff_fita[2,2],3)` (miles per gallon). The P-value of `r  round(coeff_fita[2,4],3)` (statistically significant) confirms that the __null hypothersis__ (having an increase/ decrease over the reference group null) can be rejected.

Based on the linear model previosly created we can state that 

*  __automatic__ models use an estimated mpg of `r  round(coeff_fita[1,1],3)` (miles per gallon) with a 95% confidence interval included in [`r round(cia_95_beta0[1],3)` - `r round(cia_95_beta0[2],3)`] miles per gallon.
*  __manual__ models use an increased estimated mpg of `r  round(coeff_fita[2,1],3)` (miles per gallon) over the reference group with a 95% confidence interval included in [`r round(cia_95_beta1[1],3)` - `r round(cia_95_beta1[2],3)`] miles per gallon.

Note that this model is quite limited, __`R- squared`__ indicates that this model is able to explain only __`r round(summary(fit_a)$r.squared * 100,2)`%__ of the total variability of the data.